NOPT048 Notes

### 2020-04-19 misc/study-notes/math457.tex
Finished modules

\def\lcm{\mathop {\rm lcm}}
\def\Im{\mathop {\rm Im}}
\def\Sym{\mathop {\rm Sym}}
\def\End{\mathop {\rm End}}
\def\Ann{\mathop {\rm Ann}}
\def\Tor{\mathop {\rm Tor}}
\def\Spec{\mathop {\rm Spec}}
\def\Specmax{\Spec_{\max}}
\def\Nil{\mathop {\rm Nil}}
\def\Jac{\mathop {\rm Jac}}
\def\inj{\hookrightarrow}
\def\surj{\rightarrow\!\!\!\!\!\rightarrow}
\def\defn{\noindent {\bf Definition.\enspace}}
\def\bar{\overline}
\def\mod{\!\!\!\!\pmod}

\maketitle{MATH 457 Review}{by}{Marcel K. Goh}{21 April 2020}

@@

A {\it ring} $R$ is a set with operations $+$ and $\cdot$ such that
\medskip
\item{i)} $(R,+)$ is an abelian group;
\smallskip
\item{ii)} $(R,\cdot)$ is a semigroup;
\smallskip
\item{iii)} $\cdot$ distributes over $+$ on both sides:
$$a\cdot (b+c) = a\cdot b + a\cdot c \quad\hbox{and}\quad (a+b)\cdot c = a\cdot c + b\cdot c$$
A {\it semiring is the same as a ring} except that condition (i) above becomes
\medskip
\item{i')} $(R,+)$ is a monoid with absorbing unit 0.
\medskip
A ring is {\it unital} if $(R,\cdot)$ has a unit $1$. We always assume that $1\neq 0$, since if $1=0$ then $R = \{0\}$. Observe that in a unital ring, $(R,+)$ is necessarily abelian. A ring is said to be {\it commutative} if $(R,\cdot)$ is.

@@

\proof We need to construct the section $s$ of $\pi : M\surj F$.

>>>>

\def\lcm{\mathop {\rm lcm}}
\def\Im{\mathop {\rm Im}}
\def\Id{\mathop {\rm Id}}
\def\Sym{\mathop {\rm Sym}}
\def\End{\mathop {\rm End}}
\def\Ann{\mathop {\rm Ann}}
\def\Tor{\mathop {\rm Tor}}
\def\Spec{\mathop {\rm Spec}}
\def\Specmax{\Spec_{\max}}
\def\Nil{\mathop {\rm Nil}}
\def\Jac{\mathop {\rm Jac}}
\def\inj{\hookrightarrow}
\def\surj{\rightarrow\!\!\!\!\!\rightarrow}
\def\defn{\noindent {\bf Definition.\enspace}}
\def\bar{\overline}
\def\mod{\!\!\!\!\pmod}

\maketitle{MATH 457 Honours Algebra 4\footnote{$^*$}{\eightpt Course given by Prof. Mika\"el Pichot at McGill University}}{Notes by}{Marcel K. Goh}{23 April 2020}

@@

A {\it ring} $R$ is a set with operations $+$ and $\cdot$ such that
\medskip
\item{i)} $(R,+)$ is an abelian group;
\smallskip
\item{ii)} $(R,\cdot)$ is a semigroup;
\smallskip
\item{iii)} $\cdot$ distributes over $+$ on both sides:
$$a\cdot (b+c) = a\cdot b + a\cdot c \quad\hbox{and}\quad (a+b)\cdot c = a\cdot c + b\cdot c$$
A {\it semiring is the same as a ring} except that condition (i) above becomes
\medskip
\item{i')} $(R,+)$ is a monoid with absorbing identity 0.
\medskip
A ring is {\it unital} if $(R,\cdot)$ has a unit $1$. We always assume that $1\neq 0$, since if $1=0$ then $R = \{0\}$. Observe that in a unital ring, $(R,+)$ is necessarily abelian. A ring is said to be {\it commutative} if $(R,\cdot)$ is.

@@

\proof We need to construct the section $s$ of $\pi : M\surj F$. Let $A$ be a basis of $F$. Since $\pi$ is surjective, for any $a\in A$ we can find $m_a\in M$ such that $\pi(m_a) = m$. This gives a map $s_* : A \to M$ and by the universal property there is a unique extension $s : F_A\to M$. We have $\pi \circ s = \Id$ on the basis and therefore everywhere on $F$. So $s$ is a section of $\pi$. Let $F' = \Im(s) \subseteq M$. So $F'\cong F$ as $R$-modules. We claim that $M = N\oplus F'$ (viewing $N$ as a submodule of $M$).

Firstly, $N\cap F' = \{0\}$, since if $m\in N\cap F'$, then $\pi(m) = 0$ and there exists $f\in F$ such that $s(f) = m$. But this implies that $f = \pi\big(s(f)\big) = \pi(m) = 0$, so $m=0$. And $M = N + F'$ because any $m\in M$ can be expressed as the sum of $\big(m-s\circ \pi(m)\big) + s\circ\pi (m)$.\slug

The following theorem shows that the rank of a free module is well-defined.

\proclaim Theorem R. If $R^n\cong R^m$ as $R$-modules, then $n = m$.

\proof Since $R$ is unital and commutative, it contains a maximal ideal $M$. Let $K = R/M$ and consider the submodule $MR^n = \big\{s(x_1,\cdots,x_n) \in R^n: s\in M, x_i\in R\big\}$. The quotient module is $K^n = (R/M)^n$ and a module isomorphism $R^n \cong R^m$ descends to a $R$-module isomorphism $K^n \cong K^m$. This map has kernel $M$ and is a $K$-vector space isomorphism. So the dimension of the two vector spaces are the same and thus $n=m$.\slug

Let $M$ be a module over an integral domain $R$. The set of {\it torsion elements}
$$\Tor(M) = \{m\in M : rm = 0\ \hbox{for some}\ r\neq 0\}$$
is a submodule of $M$. A module is called {\it torsion} if $\Tor(M) = M$ and {\it torsion-free} if $\Tor(M) = \{0\}$. Note that $R^n$ is torsion-free, since it has a basis $\{e_n\}$ and if $am = ra_1e_1 + \cdots + ra_ne_n = 0$, then $ra_i = 0$ for all $i$ and $m = 0$.

\proclaim Proposition T. For any module $M$ over an integral domain $R$, $M/Tor(M)$ is torsion-free.

\proof Let $N = M/\Tor(M)$ and let $\bar m\in N$. Suppose there exists $r\neq 0$ such that $r\bar m = 0$. So $\bar{rm} = 0$ and $rm \in \Tor(M)$. Thus there exists $s\neq 0$ such that $(rs)\bar m = 0$. But $\bar{rs}\neq 0$ so $m$ must be 0. Hence $\Tor(N) = \{0\}$.\slug

\proclaim Lemma G. A module $M$ over an integral domain $R$ is torsion if and only if it is generated by torsion elements.

\proof The forward direction is clear. Conversely, suppose $M = (A)$ and every element in $A$ is torsion. Let $m = s_1a_1 + \cdots + s_na_n \in M$ for some $s_i\in R$ and $a_i\in A$. Each $a_i$ is a torsion element so there is $r_i$ such that $r_ia_i = 0$. Let $r = r_1\cdots r_n$. Then $rm = 0$.\slug

\proclaim Proposition F. Let $M$ be a finitely generated module over an integral domain $R$. There exists a free module $F\subseteq M$ such that $M/F$ is torsion.

\proof Let $M = (A)$ where $A$ is finite. Let $B\subseteq A$ be a maximal basis which generates a free module $F$ of rank $n = |B|$. Let $N$ be the quotient $M/F$. For every $a\in A\setminus B$, the module $(B\cup \{a\})$ is not free. So there exists $r\in R, r_b\in R$, not all zero, such that
$$ra + \sum_{b\in B} r_b b = 0.$$
Note that $r\neq 0$, otherwise $B$ would not be a basis. But $ra = 0 \bmod F$, so $N$ is generated by torsion elements and by the previous lemma, $N$ is torsion.\slug

\section 5. Modules over PIDs

An $R$-module has properties very much like a vector space when $R$ is a PID.

\proclaim Proposition F. Let $R$ be a PID. Then every submodule of a free module $R^n$ is free of rank $k\leq n$.

\proof We proceed by induction. When $n=1$, every ideal $I\subseteq R$ is free, isomorphic to $R$. Now assume the proposition is true for $R^n$. Let $M$ be a submodule of $R^{n+1}$. Let $\pi : R^{n+1} \surj R^n$ be the projection map on to the first $n$ coordinates. So we have a short exact sequence
$$\ker(\pi_{|M}) \inj M \surj \pi(M).$$
But $\pi(M)$ is a submodule of $R^n$ so, by the induction hypothesis, it is free and the sequence splits. Thus $M\cong \ker(\pi_{|M}) \oplus \pi(M)$ is free.\slug

A module $M$ over a unital ring $R$ is called a {\it Noetherian module} if every submodule is finitely generated.

\proclaim Propostion N. Let $M$ be a left $R$-module. The following are equivalent:
\medskip
\item{i)} $M$ is Noetherian.
\smallskip
\item{ii)} $M$ satisfies the ascending chain condition on left modules.
\smallskip
\item{iii)} If $\frak F$ is a nonempty family of submodules, there exists a maximal element in $\frak F$ with respect to inclusion.
\medskip

\proof To show that (i) implies (ii), we let $N_1\subseteq N_2 \subseteq N_3 \subseteq \cdots$ be an increasing sequence of modules. We need to know there is an upper bound. Let $N = \bigcup_{i\geq 1} N_i$. Since $N$ is finitely generated, there will be a first index $i$ such that all the generators of $N$ belong to $N_i$. Thus $N = N_i$ for some $i$.

We show that (ii) implies (iii) by contraposition. If (iii) fails, then there exists a family $\frak F$ of submodules for which there is no maximal element. Pick $N_1\in \frak F$. We can find $N_2$ such that $N_1$ is properly contained in $N_2$. Continuing in this way, we are left with an increasing chain that does not stabilise.

Lastly, assume that (i) does not hold; i.e.\ there is a submodule $N$ that is not finitely generated. Let $\{n_1,n_2,\ldots\}$ be an infinite countable subset of $N$ such that, for every $k$, $N_k = (n_1,\ldots,n_k)$ is properly contained in $N_{k+1} = (n_1,\ldots,n_{k+1})$. Now $\frak F = \{N_k\}$ is a family of submodules without a maximal element, so (iii) fails.\slug

\proclaim Proposition S. Let $N \inj P\surj Q$ be an exact sequence of modules. Then $N$ and $Q$ are Noetherian if and only if $P$ is Noetherian.

\proof Clearly $N$ is Noetherian if $P$ is. To show that $Q$ is Noetherian, let $M\subseteq Q$ be a submodule. Then, by the fourth isomorphism theorem, $M$ is the image of a submodule $M'$ of $P$. Since $M'$ is finitely generated, $M$ is as well.

Conversely, assume that $N$ and $Q$ are Noetherian and let $M\subseteq P$ be a submodule. Let $\pi:P\surj Q$ be the quotient map and consider the exact sequence $\ker(\pi_{|M})\inj M\surj \pi(M)$. Let $X$ be a finite generating set for $\ker(\pi_{|M})$ and $Y$ be a finite set in $M$ such that $\bar Y = \pi(Y)$ is a finite generating set of $\pi(M)$. Then for any $m\in M$, then there exist some $r_x$ and $r_y$ in $R$ such that
$$m = \sum_{x\in X} r_x x + \sum_{y\in Y} r_y y$$
and $X\cup Y$ generates $M$.\slug

\proclaim Theorem R. The following are equivalent:
\medskip
\item{i)} $R$ is a Noetherian ring.
\smallskip
\item{ii)} The free module $R^n$ is Noetherian for every $n$.
\smallskip
\item{iii)} Every finitely generated $R$-module is Noetherian.\noskipslug
\medskip

A corollary of Theorem R is that every finitely generated module over a PID is Noetherian. For example, $R = K[x_1,\ldots,x_n]$ is Noetherian.

\proclaim Lemma N. If $R$ is a PID and $M$ a torsion-free $R$-module, then $M$ is free.

\proof In general, we showed that there exists $F$ free such that $F\inj M \surj T$, where $T$ is torsion. Since $M$ is Noetherian, we can choose a maximal $F$ satisfying this property. We claim that $M=F$. Let $\pi:M\surj T$ denote the quotient map and let $m\in M$. Since $\pi(m)\in T$ is a torsion element, there exists $r\in R$ such that $r\pi(m) = 0$. So $rm\in \ker\pi$ and $rm\in F$. Let $f_r : M \to M$ be the map that sends $m\mapsto rm$. This map is injective because $M$ is torsion-free. Since $f_r(F) \subseteq F$ and $f_r(M) \subseteq F$, so the submodule $f_r(F,M)$ is contained in $F$. By Proposition F, $f_r(F,M)$ is free, so $M$ is free.\slug

\proclaim Theorem T. Let $M$ be a finitely generated module over a PID $R$. Then $M\cong R^n \oplus \Tor(M)$.

\proof Consider the exact sequence $\Tor(M)\inj M\surj N$ where $N$ is torsion-free. Since $R$ is a PID, $N$ is free and the sequence splits, giving us the desired direct sum decomposition.\slug

The integer $n$ given by Theorem T is called the {\it free rank} of a module. If two modules $M$ and $N$ are isomorphic, then their free ranks are equal and $\Tor(M)\cong \Tor(N)$. The following theorem is called the structure theorem for finitely generated modules over a PID.

\proclaim Theorem S. Let $R$ be a PID and let $F \cong R^n$ be a finitely generated free module. Let $M$ be a finitely generated submodule of $F$. Then there exists a basis $(e_1,\ldots, e_n)$ of $F$ and elements $r_1,\ldots,r_m\in R$ such that $(r_1e_1,\ldots,r_me_m)$ forms a basis of $M$:
$$F/M \cong R^{n-m} \oplus R/(r_1) \oplus \cdots \oplus R/(r_m)$$
The elements $r_i$ are unique up to multiplication by a unit if we assume that $r_i$ divides $r_{i+1}$.\slug

The elements $r_i$ in Theorem S are called the {\it invariant factors} of the module.
